**Measuring Depression and Anxiety with 4 items? Adaptation of the PHQ-4 to increase its Sensitivity to Subclinical Variability**

### 2024-12-10 paper/4_WIP/manuscript.qmd
add paragraph

  The PHQ-4 is an ultra-brief (4 items) screening questionnaire for depression and anxiety. In this brief report, we test the benefits of adding one additional response option ("Once or twice", in between "Not at all" and "Several days") to improve the scale's sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. In study 1 (n=485), we provide evidence using Item Response Theory (IRT) that the new response option does improve the scale's psychometric quality and extends the sensitivity to the measured constructs on the lower end of the spectrum. In study 2 (n=836), we show that the refined version offers an improved sensitivity to subclinical variability in depression (indexed by the BDI-II) as compared to the original version. In conclusion, adding the "once or twice" response option is a low-cost no-downsides way of increasing the PHQ-4's sensitivity to subclinical variability, making it a tool of choice for general population research.

@@

Item Response Theory (IRT) provides insights into how well items and responses capture an underlying latent trait $\theta$. For each of the subscales, we fitted a unidimensional graded response model [GRM, @samejima1997graded]. For anxiety, the two items captured 89.2\% of the variance of the latent anxiety dimension ($\theta_{anxiety}$). The discrimination parameters suggested that the first item was less precise ($\alpha = 3.42$) than the second item ($\alpha = 12.55$) in its ability to discriminate between various levels of anxiety (i.e., each response on the second item covers a more exclusive range of $\theta_{anxiety}$, as can be seen in **Figure 1B**). The two depression items captured 82.8\% of the variance of its latent trait ($\theta_{depression}$), and the opposite pattern was found: the first item had a higher precision ($\alpha = 16.46$) than the first ($\alpha = 2.41$). However, it is important to note that the "less precise" items were also the ones covering a larger portion of the latent space (being more sensitive especially on the lower end of the spectrum), offering an interesting trade-off between sensitivity and precision. Importantly for our objective, the added "Once or twice" option did cover a selective and unique portion of the latent space.

@@

>>>>

  The PHQ-4 is an ultra-brief (4 items) screening questionnaire for depression and anxiety. In this brief report, we test the benefits of adding one additional response option ("Once or twice", in between "Not at all" and "Several days") to improve the scale's sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. In study 1 (N=485), we provide evidence using Item Response Theory (IRT) that the new response option does improve the scale's psychometric quality and extends the sensitivity to the measured constructs on the lower end of the spectrum. In study 2 (N=836), we show that the refined version offers an improved sensitivity to subclinical variability in depression (indexed by the BDI-II) as compared to the original version. In conclusion, adding the "once or twice" response option is a low-cost no-downsides way of increasing the PHQ-4's sensitivity to subclinical variability, making it a tool of choice for general population research.

@@

Item Response Theory (IRT) provides insights into how well items and responses capture an underlying latent trait $\theta$. For each of the subscales, we fitted a unidimensional graded response model [GRM, @samejima1997graded]. For anxiety, the two items captured 89.2\% of the variance of the latent anxiety dimension ($\theta_{anxiety}$). The discrimination parameters suggested that the first item was less precise ($\alpha = 3.42$) than the second item ($\alpha = 12.55$) in its ability to discriminate between various levels of anxiety (i.e., each response on the second item covers a more exclusive range of $\theta_{anxiety}$, as can be seen in @fig-one). The two depression items captured 82.8\% of the variance of its latent trait ($\theta_{depression}$), and the opposite pattern was found: the first item had a higher precision ($\alpha = 16.46$) than the first ($\alpha = 2.41$). However, it is important to note that the "less precise" items were also the ones covering a larger portion of the latent space (being more sensitive especially on the lower end of the spectrum), offering an interesting trade-off between sensitivity and precision. Importantly for our objective, the added "Once or twice" option did cover a selective and unique portion of the latent space.

@@

One natural methodological limitation pertains to the interpretation of the latent dimensions in an IRT framework applied to pairs of items. In this study's context, references to for instance "the latent anxiety dimension" merely corresponds to the amalgamation of the two items of the anxiety subscale, and not to a more general and valid true anxiety factor.
